---
title: 'Impact of Speeding and Vehicle Types on Crash Severity'
author: "Team Members: Shivansh Patel, Manas Mathur, Muhammad Saleem"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Research Topic

The research project aims to analyze the **impact of speeding and vehicle types on crash severity**. We are interested in understanding whether there is a significant relationship between different vehicle types, speeding incidents, and the severity of crashes that occur on the roads. Additionally, we will compare the **crash fatality rate of U.S. crashes** with crashes on the **Autobahn**, where there are segments with no speed limit, to understand the impact of speed regulations on crash outcomes.

## Team Members

- Shivansh Patel
- Manas Mathur
- Muhammad Saleem

## Data Description and Cleaning Steps

For our analysis, we will use multiple datasets obtained from the following sources:

- **U.S. Crash Data**: Data obtained from [catalog.data.gov](https://catalog.data.gov/dataset/crash-reporting-drivers-data/resource/9851a37f-4f32-464e-8ba6-c23023653a7f) and other sources like the [Insurance Institute for Highway Safety (IIHS)](https://www.iihs.org/topics/fatality-statistics/detail/yearly-snapshot) and the [National Highway Traffic Safety Administration (NHTSA)](https://www.nhtsa.gov/file-downloads?p=nhtsa/downloads/CISS/). This dataset (`US.csv`) contains detailed information on crashes that occurred in the United States, including vehicle types, speeding status, crash severity, and weather conditions.

- **Autobahn Crash Data**: Data obtained from an **Excel dataset titled `AutoBahn.xls`** that contains crash information for the Autobahn, where some sections do not have speed limits. The dataset includes details on crash fatalities, vehicle types, and factors influencing crash severity.

### Loading the Data

First, we will load both datasets into R for further analysis.

```{r load_data, warning=FALSE, message=FALSE}
# Load necessary packages
library(dplyr)
library(ggplot2)
library(readxl)

# Load the U.S. crash data
#us_crash_data <- read.csv("US.csv")

# Load the Autobahn crash data
#autobahn_data <- read_excel("AutoBahn.xls")

# View the structure of both datasets
#str(us_crash_data)
#str(autobahn_data)
```

### Initial Data Cleaning

We will start by cleaning both datasets to remove missing or incorrect data entries, harmonize variable names, and prepare them for comparison.

```{r data_cleaning, warning=FALSE, message=FALSE}
# Cleaning U.S. crash data
#us_crash_data_clean <- us_crash_data

# Cleaning Autobahn crash data
#autobahn_data_clean <- autobahn_data

# Summary of cleaned datasets
#summary(us_crash_data_clean)
#summary(autobahn_data_clean)
```

### Marginal Summaries

After cleaning the data, we generate marginal summaries to understand the distribution of key variables, such as **crash severity**, **vehicle type**, and **speeding involvement** in both datasets.

```{r marginal_summaries, warning=FALSE, message=FALSE}
# Marginal summaries of U.S. crash data
#summary(us_crash_data_clean$crash_severity)
#table(us_crash_data_clean$vehicle_type)
#table(us_crash_data_clean$speeding)

# Marginal summaries of Autobahn crash data
#summary(autobahn_data_clean$Crash_Severity)
#table(autobahn_data_clean$Vehicle_Type)
#table(autobahn_data_clean$Fatality)
```

## Questions to Be Addressed

### Fleshed Out Project Idea

The primary questions we aim to address in this project are:

1. **How does speeding impact the severity of crashes on U.S. roads compared to the Autobahn?**
   - We will investigate whether crashes involving speeding tend to have higher severity in the U.S. compared to the Autobahn, where some segments have no speed limit.

2. **Is there a significant difference in crash fatality rates between U.S. crashes and Autobahn crashes?**
   - We will compare the fatality rates of crashes on U.S. roads versus those on the Autobahn to understand the impact of speed limits on crash outcomes.

3. **Do certain vehicle types result in more severe crashes on U.S. roads compared to the Autobahn?**
   - We will examine whether certain vehicle types (e.g., motorcycles, SUVs, trucks) are more prone to severe crashes and compare the results between the U.S. and the Autobahn.

To answer these questions, we will use data visualizations such as **bar charts**, **histograms**, and **scatterplots**, along with **statistical modeling** where appropriate. We also plan to consider possible confounding variables such as weather conditions, road type, and driver behavior.
