---
title: 'Analysis of Traffic Crash Patterns and Their Contributing Factors'
author: "Shivansh Patel, Manas Mathur, Muhammad Raham Saleem"
subtitle: 'DS2020 - Final Project'
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

This analysis explores traffic crash patterns in our dataset, focusing on key factors that influence crash occurrence and severity. We investigate the relationships between speed limits, vehicle types, substance abuse, and various other factors that contribute to traffic accidents.

## Data Preparation

```{r load-libraries}
# Load required libraries
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(corrplot)
library(forcats)
library(scales)

# Load the dataset
crash_data <- read.csv("US.csv")

# Basic text cleaning function
clean_text_column <- function(x) {
  x <- toupper(trimws(x))
  x[x == ""] <- "UNKNOWN"
  x[is.na(x)] <- "UNKNOWN"
  return(x)
}

# Clean and standardize the data
crash_clean <- crash_data %>%
  # First clean all character columns
  mutate(across(where(is.character), clean_text_column)) %>%
  # Then handle specific columns
  mutate(
    Weather = case_when(
      Weather == "N/A" ~ "UNKNOWN",
      grepl("FREEZING RAIN|FREEZING DRIZZLE", Weather) ~ "FREEZING PRECIPITATION",
      grepl("BLOWING", Weather) ~ "BLOWING WEATHER",
      TRUE ~ Weather
    ),
    Traffic.Control = case_when(
      Traffic.Control == "N/A" ~ "UNKNOWN",
      grepl("SIGNAL", Traffic.Control) ~ "TRAFFIC SIGNAL",
      TRUE ~ Traffic.Control
    ),
    Driver.Substance.Abuse = case_when(
      Driver.Substance.Abuse == "N/A" ~ "UNKNOWN",
      Driver.Substance.Abuse == "NONE DETECTED" ~ "NONE",
      TRUE ~ Driver.Substance.Abuse
    ),
    Vehicle.Body.Type = case_when(
      Vehicle.Body.Type == "N/A" ~ "UNKNOWN",
      grepl("PASSENGER", Vehicle.Body.Type) ~ "PASSENGER VEHICLE",
      TRUE ~ Vehicle.Body.Type
    )
  )

# Handle rare categories directly
crash_clean <- crash_clean %>%
  # Weather categories
  group_by(Weather) %>%
  mutate(count = n()) %>%
  ungroup() %>%
  mutate(Weather = if_else(count < nrow(crash_clean) * 0.001, "OTHER", Weather)) %>%
  select(-count) %>%
  
  # Traffic Control categories
  group_by(Traffic.Control) %>%
  mutate(count = n()) %>%
  ungroup() %>%
  mutate(Traffic.Control = if_else(count < nrow(crash_clean) * 0.001, "OTHER", Traffic.Control)) %>%
  select(-count) %>%
  
  # Vehicle Body Type categories
  group_by(Vehicle.Body.Type) %>%
  mutate(count = n()) %>%
  ungroup() %>%
  mutate(Vehicle.Body.Type = if_else(count < nrow(crash_clean) * 0.001, "OTHER", Vehicle.Body.Type)) %>%
  select(-count)

# Create severity score
crash_severity <- crash_clean %>%
  mutate(
    severity_score = case_when(
      Injury.Severity == "FATAL INJURY" ~ 4,
      Injury.Severity == "SUSPECTED SERIOUS INJURY" ~ 3,
      Injury.Severity == "SUSPECTED MINOR INJURY" ~ 2,
      Injury.Severity == "POSSIBLE INJURY" ~ 1,
      Injury.Severity == "NO APPARENT INJURY" ~ 0,
      TRUE ~ NA_real_
    )
  )

# Print summary to verify the cleaning
cat("Unique values in Weather after cleaning:\n")
print(table(crash_clean$Weather))
cat("\nUnique values in Traffic.Control after cleaning:\n")
print(table(crash_clean$Traffic.Control))
cat("\nUnique values in Vehicle.Body.Type after cleaning:\n")
print(table(crash_clean$Vehicle.Body.Type))
```

## Analysis 1: Speed Zones and Crash Patterns

### Research Question
What speed zones result in the most crashes, and how does this affect crash severity

```{r speed-analysis}


colnames(crash_clean)
unique(crash_clean$Driver.Distracted.By)


speed_limit_crashes <- table(crash_clean$Speed.Limit)
speed_limit_summary <- data.frame(
    speed_limit = names(speed_limit_crashes),
    crash_count = as.numeric(speed_limit_crashes)
)


speed_limit_summary <- speed_limit_summary[order(-speed_limit_summary$crash_count), ]


print(speed_limit_summary)


library(ggplot2)
ggplot(speed_limit_summary, aes(x = speed_limit, y = crash_count)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    theme_minimal() +
    labs(
        title = "Crash Frequency by Speed Limit",
        x = "Speed Limit",
        y = "Number of Crashes"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

head(crash_clean$Crash.Date.Time, 1)

nrow(crash_clean)

```
## Analysis 2: Lights, surfaces, and injury severity

### Research Question
What combinations of lighting and surfaces result in the worst injuries?
```{r}

crash_analysis <- crash_clean %>%
  
  filter(!Light %in% c("N/A", 
                      "DARK -- UNKNOWN LIGHTING", 
                      "UNKNOWN", 
                      "OTHER", 
                      "DARK - UNKNOWN LIGHTING")) %>%
  filter(!Surface.Condition %in% c("N/A", 
                                 "OTHER", 
                                 "UNKNOWN")) %>%

  mutate(Light = case_when(
    Light %in% c("DARK LIGHTS ON", "DARK - LIGHTED") ~ "DARK WITH LIGHTS",
    Light %in% c("DARK NO LIGHTS", "DARK - NOT LIGHTED") ~ "DARK WITHOUT LIGHTS",
    TRUE ~ Light
  )) %>%

  mutate(Surface.Condition = case_when(
    Surface.Condition %in% c("WATER(STANDING/MOVING)", "WATER (STANDING, MOVING)") ~ "WATER (STANDING/MOVING)",
    TRUE ~ Surface.Condition
  )) %>%
  mutate(Surface.Condition = case_when(
    Surface.Condition %in% c("ICE/FROST", "ICE") ~ "ICE",
    TRUE ~ Surface.Condition
  )) %>%

  mutate(
    severity_score = case_when(
      Injury.Severity == "FATAL INJURY" ~ 4,
      Injury.Severity == "SUSPECTED SERIOUS INJURY" ~ 3,
      Injury.Severity == "SUSPECTED MINOR INJURY" ~ 2,
      Injury.Severity == "POSSIBLE INJURY" ~ 1,
      Injury.Severity == "NO APPARENT INJURY" ~ 0,
      TRUE ~ NA_real_
    )
  )


severity_by_conditions <- crash_analysis %>%
  group_by(Light, Surface.Condition) %>%
  summarise(
    avg_severity = mean(severity_score, na.rm = TRUE),
    n_crashes = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(avg_severity))


print(head(severity_by_conditions, 10))

ggplot(severity_by_conditions, 
       aes(x = Surface.Condition, 
           y = avg_severity,
           fill = Light)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  coord_flip() +  
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10),
    panel.grid.major.y = element_blank(), 
    legend.position = "bottom"
  ) +
  labs(title = "Average Crash Severity by Light and Surface Conditions",
       subtitle = "Severity Score: 0 (No Injury) to 1 (Fatal)",
       x = "Surface Condition",
       y = "Average Severity Score",
       fill = "Lighting Condition")

```
```{r}

substance_present <- c("ALCOHOL CONTRIBUTED", "ALCOHOL PRESENT",
                      "COMBINATION CONTRIBUTED", "COMBINED SUBSTANCE PRESENT",
                      "ILLEGAL DRUG CONTRIBUTED", "ILLEGAL DRUG PRESENT",
                      "MEDICATION CONTRIBUTED", "MEDICATION PRESENT",
                      "SUSPECT OF ALCOHOL USE, SUSPECT OF DRUG USE",
                      "NOT SUSPECT OF ALCOHOL USE, SUSPECT OF DRUG USE",
                      "SUSPECT OF ALCOHOL USE, NOT SUSPECT OF DRUG USE")


crash_analysis <- crash_clean %>%
  mutate(
    
    crash_datetime = as.POSIXct(Crash.Date.Time, format = "%m/%d/%Y %I:%M:%S %p"),
    
    day_of_week = weekdays(crash_datetime),
    
    day_of_week = factor(day_of_week, 
                        levels = c("Monday", "Tuesday", "Wednesday", 
                                 "Thursday", "Friday", "Saturday", "Sunday")),
    
    severity_score = case_when(
      Injury.Severity == "FATAL INJURY" ~ 4,
      Injury.Severity == "SUSPECTED SERIOUS INJURY" ~ 3,
      Injury.Severity == "SUSPECTED MINOR INJURY" ~ 2,
      Injury.Severity == "POSSIBLE INJURY" ~ 1,
      Injury.Severity == "NO APPARENT INJURY" ~ 0,
      TRUE ~ NA_real_
    ),

    has_substance = Driver.Substance.Abuse %in% substance_present
  )


severity_by_day <- crash_analysis %>%
  group_by(day_of_week, has_substance) %>%
  summarise(
    avg_severity = mean(severity_score, na.rm = TRUE),
    n_crashes = n(),
    se = sd(severity_score, na.rm = TRUE) / sqrt(n()),  
    .groups = 'drop'
  )

ggplot(severity_by_day) +
  geom_line(aes(x = day_of_week, 
                y = avg_severity, 
                group = has_substance,
                color = has_substance),
            size = 1.2) +
  geom_point(aes(x = day_of_week, 
                 y = avg_severity,
                 color = has_substance),
             size = 3) +

  scale_color_manual(values = c("lightblue", "red"),
                    labels = c("Non-Substance Related", "Substance Related")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    title = "Average Crash Severity by Day of Week",
    subtitle = "Comparing Substance-Related vs Non-Substance-Related Crashes",
    x = "Day of Week",
    y = "Average Severity Score (0-4)"
  )
```




## Analysis 3: Impact Locations and Crash Patterns

### Research Question
How do Impact Zones affect crash severity


```{r}
library(sf)
library(tigris)
library(dplyr)
library(lubridate)

crash_coords <- crash_clean %>% 
  select(Longitude, Latitude)

nrow(crash_coords)

crash_sf <- st_as_sf(crash_coords, coords = c("Longitude", "Latitude"), crs = 4326)

states <- states(cb = TRUE) %>%
  st_transform(crs = 4326)  

crash_with_states <- st_join(crash_sf, states["NAME"])

crash_states_df <- crash_with_states %>%
  mutate(row_num = row_number()) %>%
  st_drop_geometry() %>%
  select(NAME, row_num)

crash_clean_with_states <- crash_clean %>%
  mutate(row_num = row_number()) %>%
  left_join(crash_states_df, by = "row_num") %>%
  rename(state = NAME) %>%
  select(-row_num)

crash_maryland_injuries <- crash_clean_with_states %>% 
  mutate(date = mdy_hms(Crash.Date.Time)) %>%
  filter(year(date) == 2024,
#         state == "Maryland", 
         Injury.Severity != "NO APPARENT INJURY",
#         Injury.Severity != "POSSIBLE INJURY",
         Injury.Severity != "UNKNOWN",
         Vehicle.First.Impact.Location != "NON-COLLISION",
         Vehicle.First.Impact.Location != "VEHICLE NOT AT SCENE")

colnames(crash_maryland_injuries)

```

```{r}
colnames(crash_maryland_injuries)

unique(crash_maryland_injuries$Injury.Severity)

crash_maryland_injuries$Injury.Severity <- factor(crash_maryland_injuries$Injury.Severity, levels = c('SUSPECTED MINOR INJURY', 'SUSPECTED SERIOUS INJURY', 'FATAL INJURY'))

impact_injury_table <- table(crash_maryland_injuries$Vehicle.First.Impact.Location, 
                              crash_maryland_injuries$Injury.Severity)

impact_injury_prop <- prop.table(impact_injury_table, margin = 1) * 100




ggplot(data = as.data.frame.table(impact_injury_prop), 
       aes(x = reorder(Var1, -table(Var1)), y = Freq, fill = Var2)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_minimal() +
  labs(x = "Impact Location", 
       y = "Count",
       fill = "Injury Severity",
       title = "Relationship between Impact Location and Injury Severity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r}
chisq_test <- chisq.test(table(crash_maryland_injuries$Vehicle.First.Impact.Location, 
                              crash_maryland_injuries$Injury.Severity))

chisq_test
library(effectsize)
cramers_v <- cramers_v(crash_maryland_injuries$Vehicle.First.Impact.Location, 
                       crash_maryland_injuries$Injury.Severity)

stdres <- chisq_test$stdres
stdres
```
```{r}
impact_freq <- table(crash_maryland_injuries$Vehicle.First.Impact.Location)
impact_freq_sorted <- sort(impact_freq, decreasing = TRUE)

impact_chisq <- chisq.test(impact_freq)

library(ggplot2)
impact_df <- data.frame(
  location = names(impact_freq),
  count = as.numeric(impact_freq)
)

ggplot(impact_df, aes(x = reorder(location, -count), y = count)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Impact Location",
       y = "Frequency",
       title = "Distribution of Vehicle Impact Locations")
```

```{r}
weights <- 1 / table(crash_maryland_injuries$Vehicle.First.Impact.Location)
weights <- weights[crash_maryland_injuries$Vehicle.First.Impact.Location]

weighted.chisq <- chisq.test(table(crash_maryland_injuries$Vehicle.First.Impact.Location, 
                                  crash_maryland_injuries$Injury.Severity),
                            weights)
weighted.chisq
```
```{r}



first_impact_severity <- crash_severity %>%
  group_by(Vehicle.First.Impact.Location) %>%
  summarise(
    avg_severity = mean(severity_score, na.rm = TRUE),
    crash_count = n()
  ) %>%  filter(Vehicle.First.Impact.Location != "UNKNOWN", Vehicle.First.Impact.Location != "VEHICLE NOT AT SCENE",crash_count > 10)

ggplot(first_impact_severity, aes(x = reorder(Vehicle.First.Impact.Location, -avg_severity), y = avg_severity)) +
  geom_bar(stat = "identity", fill = "black") +
  geom_text(aes(label = round(avg_severity, 2)), vjust = -0.5) +
  labs(
    title = "Average Crash Severity by Vehicle Impact Location",
    x = "Type of impact",
    y = "Average Severity Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



```{r}
library(mapview)


crash_maryland_sf <- st_as_sf(crash_maryland_injuries, 
                            coords = c("Longitude", "Latitude"), 
                            crs = 4326)

colnames(crash_maryland_sf)

severity_colors <- c("FATAL" = "red",
                    "NO APPARENT INJURY" = "green",
                    "POSSIBLE INJURY" = "blue",
                    "SUSPECTED MINOR INJURY" = "yellow",
                    "SUSPECTED SERIOUS INJURY" = "orange"
                    )




mapview(crash_maryland_sf,
        zcol = "Injury.Severity",
        col.regions = severity_colors,
        cex = 3,
        alpha = 0.8,
        legend = TRUE,
        popup = c("Injury.Severity"),
        layer.name = "Crash Severity")

mapview(crash_maryland_sf,
        zcol = "Vehicle.First.Impact.Location",
        col.regions = severity_colors,
        cex = 3,
        alpha = 0.8,
        legend = TRUE,
        popup = c("Injury.Severity"),
        layer.name = "Crash Severity")



```

## Analysis 2: Factors Affecting Crash Severity

### Research Question
What factors correlate most strongly with crash severity?

```{r time-analysis}
# Extract hour from Crash.Date.Time
crash_severity <- crash_severity %>%
  mutate(
    crash_hour = hour(as.POSIXct(Crash.Date.Time, format="%m/%d/%Y %I:%M:%S %p")),
    time_of_day = case_when(
      crash_hour >= 5 & crash_hour < 12 ~ "Morning",
      crash_hour >= 12 & crash_hour < 17 ~ "Afternoon",
      crash_hour >= 17 & crash_hour < 22 ~ "Evening",
      TRUE ~ "Night"
    )
  )

# Analyze severity by time of day
time_severity <- crash_severity %>%
  group_by(time_of_day) %>%
  summarise(
    avg_severity = mean(severity_score, na.rm = TRUE),
    crash_count = n()
  )

# Visualize time of day patterns
ggplot(time_severity, aes(x = time_of_day, y = avg_severity)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = round(avg_severity, 2)), vjust = -0.5) +
  labs(
    title = "Average Crash Severity by Time of Day",
    x = "Time of Day",
    y = "Average Severity Score"
  ) +
  theme_minimal()
```

```{r weather-analysis}
# Analyze severity by weather conditions
weather_severity <- crash_severity %>%
  group_by(Weather) %>%
  summarise(
    avg_severity = mean(severity_score, na.rm = TRUE),
    crash_count = n()
  ) %>%
  filter(!is.na(Weather), Weather != "", crash_count > 10)  # Remove sparse categories

# Visualize weather patterns
ggplot(weather_severity, aes(x = reorder(Weather, -avg_severity), y = avg_severity)) +
  geom_bar(stat = "identity", fill = "darkred") +
  geom_text(aes(label = round(avg_severity, 2)), vjust = -0.5) +
  labs(
    title = "Average Crash Severity by Weather Condition",
    x = "Weather Condition",
    y = "Average Severity Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r traffic-control-analysis}
# Analyze severity by traffic control presence
control_severity <- crash_severity %>%
  group_by(Traffic.Control) %>%
  summarise(
    avg_severity = mean(severity_score, na.rm = TRUE),
    crash_count = n()
  ) %>%
  filter(!is.na(Traffic.Control), Traffic.Control != "", crash_count > 10)

# Visualize traffic control patterns
ggplot(control_severity, aes(x = reorder(Traffic.Control, -avg_severity), y = avg_severity)) +
  geom_bar(stat = "identity", fill = "forestgreen") +
  geom_text(aes(label = round(avg_severity, 2)), vjust = -0.5) +
  labs(
    title = "Average Crash Severity by Traffic Control Type",
    x = "Traffic Control",
    y = "Average Severity Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r distraction-analysis}
# Analyze severity by driver distraction
distraction_severity <- crash_severity %>%
  group_by(Driver.Distracted.By) %>%
  summarise(
    avg_severity = mean(severity_score, na.rm = TRUE),
    crash_count = n()
  ) %>%
  filter(!is.na(Driver.Distracted.By), Driver.Distracted.By != "", crash_count > 10)

# Visualize distraction patterns
ggplot(distraction_severity, aes(x = reorder(Driver.Distracted.By, -avg_severity), y = avg_severity)) +
  geom_bar(stat = "identity", fill = "purple") +
  geom_text(aes(label = round(avg_severity, 2)), vjust = -0.5) +
  labs(
    title = "Average Crash Severity by Driver Distraction",
    x = "Distraction Type",
    y = "Average Severity Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r movement-analysis}
# Analyze severity by vehicle movement
movement_severity <- crash_severity %>%
  group_by(Vehicle.Movement) %>%
  summarise(
    avg_severity = mean(severity_score, na.rm = TRUE),
    crash_count = n()
  ) %>%
  filter(!is.na(Vehicle.Movement), Vehicle.Movement != "", crash_count > 10)

# Visualize movement patterns
ggplot(movement_severity, aes(x = reorder(Vehicle.Movement, -avg_severity), y = avg_severity)) +
  geom_bar(stat = "identity", fill = "orange") +
  geom_text(aes(label = round(avg_severity, 2)), vjust = -0.5) +
  labs(
    title = "Average Crash Severity by Vehicle Movement",
    x = "Vehicle Movement",
    y = "Average Severity Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r statistical-tests}
# Chi-square test for categorical variables
# Weather and Severity
weather_chi <- table(crash_severity$Weather, crash_severity$Injury.Severity)
weather_test <- chisq.test(weather_chi)

# Traffic Control and Severity
control_chi <- table(crash_severity$Traffic.Control, crash_severity$Injury.Severity)
control_test <- chisq.test(control_chi)

# Create a summary of statistical tests
test_results <- data.frame(
  Factor = c("Weather", "Traffic Control"),
  Chi_Square = c(weather_test$statistic, control_test$statistic),
  P_Value = c(weather_test$p.value, control_test$p.value)
)

# Print test results
print(test_results)

# Perform logistic regression for severe crashes
crash_severity$is_severe <- crash_severity$severity_score >= 3

# Create logistic regression model
severe_model <- glm(is_severe ~ 
                     Weather + 
                     time_of_day + 
                     Traffic.Control + 
                     Vehicle.Movement,
                   data = crash_severity,
                   family = "binomial")

# Print model summary
summary(severe_model)
```

```{r create-summary-table}
# Create a summary table of the most significant factors
factor_summary <- data.frame(
  Factor = c("Time of Day", "Weather", "Traffic Control", 
             "Driver Distraction", "Vehicle Movement"),
  Average_Severity = c(
    max(time_severity$avg_severity),
    max(weather_severity$avg_severity),
    max(control_severity$avg_severity),
    max(distraction_severity$avg_severity),
    max(movement_severity$avg_severity)
  ),
  Most_Severe_Category = c(
    time_severity$time_of_day[which.max(time_severity$avg_severity)],
    weather_severity$Weather[which.max(weather_severity$avg_severity)],
    control_severity$Traffic.Control[which.max(control_severity$avg_severity)],
    distraction_severity$Driver.Distracted.By[which.max(distraction_severity$avg_severity)],
    movement_severity$Vehicle.Movement[which.max(movement_severity$avg_severity)]
  )
)

# Print the summary table
print(factor_summary)
```

## Analysis 4: Substance Abuse and Injury Correlation

### Research Question
Is there a higher injury rate correlated with substance abuse?

```{r substance-analysis}
# Analysis of substance abuse and injury severity
substance_summary <- crash_clean %>%
  group_by(Driver.Substance.Abuse) %>%
  summarise(
    crash_count = n(),
    severe_crashes = sum(Injury.Severity %in% c("FATAL INJURY", "SUSPECTED SERIOUS INJURY")),
    severity_rate = severe_crashes / crash_count
  )

# Visualization code will go here
```



## Conclusions

Our comprehensive analysis of traffic crash data has revealed critical patterns in crash severity influenced by multiple factors. The statistical evidence demonstrates that crash severity is not randomly distributed but rather follows distinct patterns tied to specific conditions and behaviors. The most severe crashes are predominantly associated with vehicle movement issues, particularly skidding incidents (0.435 severity score), followed by distracted driving behaviors, especially texting while driving (0.388 severity score). Traffic control infrastructure, specifically at signalized intersections (0.311 severity score), also plays a significant role in crash severity.
The temporal analysis revealed that nighttime driving consistently presents higher risks (0.295 severity score), while weather conditions, particularly wintry mix situations (0.299 severity score), create hazardous driving environments. These findings are supported by robust statistical evidence, with chi-square tests showing extremely significant relationships between both weather (χ² = 732.51, p < 0.001) and traffic control (χ² = 1173.62, p < 0.001) factors and crash severity.
The interplay between these factors suggests that crash severity is often the result of multiple concurrent risk factors, requiring a multi-faceted approach to traffic safety improvements. The data clearly indicates that human factors, particularly attention and vehicle control, combine with environmental and infrastructure conditions to create varying levels of risk on our roadways.
## Recommendations

Based on our analysis, we propose the following evidence-based recommendations:
1. Vehicle Control and Driver Training
The high severity of skidding-related crashes (0.435) suggests an urgent need for:

Implementation of mandatory winter driving training programs
Enhanced driver education focusing on vehicle control in adverse conditions
Promotion of vehicle safety features that assist with stability control
Regular vehicle maintenance campaigns focusing on tire and brake condition

2. Distracted Driving Prevention
With phone-related distractions showing severe outcomes (0.388), we recommend:

Strengthening enforcement of texting-while-driving laws
Implementing public awareness campaigns highlighting the statistical risks
Encouraging the use of phone-blocking technology while driving
Creating designated safe zones for necessary phone use at regular intervals on major highways

3. Infrastructure Improvements
Given the high severity at traffic signals (0.311), we suggest:

Reviewing and optimizing traffic signal timing at high-risk intersections
Installing advanced warning systems for upcoming traffic signals
Implementing smart traffic control systems that adapt to real-time conditions
Enhanced lighting and visibility improvements at signalized intersections

4. Weather-Related Safety Measures
To address weather-related severity (0.299), we propose:

Developing real-time weather alert systems for drivers
Installing dynamic speed limit signs that adjust to weather conditions
Improving road surface treatments during winter conditions
Creating weather-specific traffic management protocols

5. Nighttime Safety Enhancements
To mitigate nighttime risks (0.295), we recommend:

Upgrading street lighting in high-risk areas
Installing reflective road markers and enhanced signage
Implementing night-specific speed limits in high-risk areas
Encouraging the use of enhanced vehicle lighting systems

6. Long-Term Monitoring and Evaluation
To ensure ongoing improvement:

Establish a continuous monitoring system for crash severity patterns
Conduct regular evaluations of implemented safety measures
Update safety protocols based on new data and findings
Maintain a public dashboard of safety metrics and improvements

These recommendations should be implemented in phases, prioritizing measures that address the most severe risk factors first. Regular evaluation of their effectiveness will allow for adjustments and improvements as needed.
## References

Maryland Crash Records
