---
title: 'Analysis of Traffic Crash Patterns and Their Contributing Factors'
author: "Shivansh Patel, Manas Mathur, Muhammad Raham Saleem"
subtitle: 'DS2020 - Final Project'
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

This analysis explores traffic crash patterns in our dataset, focusing on key factors that influence crash occurrence and severity. We investigate the relationships between speed limits, vehicle types, substance abuse, and various other factors that contribute to traffic accidents.

## Data Preparation

```{r load-libraries}
# Load required libraries
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(corrplot)
library(forcats)
library(scales)

# Load the dataset
crash_data <- read.csv("US.csv")

# Basic text cleaning function
clean_text_column <- function(x) {
  x <- toupper(trimws(x))
  x[x == ""] <- "UNKNOWN"
  x[is.na(x)] <- "UNKNOWN"
  return(x)
}

# Clean and standardize the data
crash_clean <- crash_data %>%
  # First clean all character columns
  mutate(across(where(is.character), clean_text_column)) %>%
  # Then handle specific columns
  mutate(
    Weather = case_when(
      Weather == "N/A" ~ "UNKNOWN",
      grepl("FREEZING RAIN|FREEZING DRIZZLE", Weather) ~ "FREEZING PRECIPITATION",
      grepl("BLOWING", Weather) ~ "BLOWING WEATHER",
      TRUE ~ Weather
    ),
    Traffic.Control = case_when(
      Traffic.Control == "N/A" ~ "UNKNOWN",
      grepl("SIGNAL", Traffic.Control) ~ "TRAFFIC SIGNAL",
      TRUE ~ Traffic.Control
    ),
    Driver.Substance.Abuse = case_when(
      Driver.Substance.Abuse == "N/A" ~ "UNKNOWN",
      Driver.Substance.Abuse == "NONE DETECTED" ~ "NONE",
      TRUE ~ Driver.Substance.Abuse
    ),
    Vehicle.Body.Type = case_when(
      Vehicle.Body.Type == "N/A" ~ "UNKNOWN",
      grepl("PASSENGER", Vehicle.Body.Type) ~ "PASSENGER VEHICLE",
      TRUE ~ Vehicle.Body.Type
    )
  )

# Handle rare categories directly
crash_clean <- crash_clean %>%
  # Weather categories
  group_by(Weather) %>%
  mutate(count = n()) %>%
  ungroup() %>%
  mutate(Weather = if_else(count < nrow(crash_clean) * 0.001, "OTHER", Weather)) %>%
  select(-count) %>%
  
  # Traffic Control categories
  group_by(Traffic.Control) %>%
  mutate(count = n()) %>%
  ungroup() %>%
  mutate(Traffic.Control = if_else(count < nrow(crash_clean) * 0.001, "OTHER", Traffic.Control)) %>%
  select(-count) %>%
  
  # Vehicle Body Type categories
  group_by(Vehicle.Body.Type) %>%
  mutate(count = n()) %>%
  ungroup() %>%
  mutate(Vehicle.Body.Type = if_else(count < nrow(crash_clean) * 0.001, "OTHER", Vehicle.Body.Type)) %>%
  select(-count)

# Create severity score
crash_severity <- crash_clean %>%
  mutate(
    severity_score = case_when(
      Injury.Severity == "FATAL INJURY" ~ 4,
      Injury.Severity == "SUSPECTED SERIOUS INJURY" ~ 3,
      Injury.Severity == "SUSPECTED MINOR INJURY" ~ 2,
      Injury.Severity == "POSSIBLE INJURY" ~ 1,
      Injury.Severity == "NO APPARENT INJURY" ~ 0,
      TRUE ~ NA_real_
    )
  )

# Print summary to verify the cleaning
cat("Unique values in Weather after cleaning:\n")
print(table(crash_clean$Weather))
cat("\nUnique values in Traffic.Control after cleaning:\n")
print(table(crash_clean$Traffic.Control))
cat("\nUnique values in Vehicle.Body.Type after cleaning:\n")
print(table(crash_clean$Vehicle.Body.Type))
```

## Analysis 1: Speed Zones and Crash Patterns

### Research Question
What speed zones result in the most crashes, and how does this affect crash severity

```{r speed-analysis}
# Analysis of crashes by speed limit

colnames(crash_clean)
unique(crash_clean$Driver.Distracted.By)

# Assuming your data frame is called 'crash_data'
speed_limit_crashes <- table(crash_clean$Speed.Limit)
speed_limit_summary <- data.frame(
    speed_limit = names(speed_limit_crashes),
    crash_count = as.numeric(speed_limit_crashes)
)

# Sort by crash count in descending order
speed_limit_summary <- speed_limit_summary[order(-speed_limit_summary$crash_count), ]

# View the results
print(speed_limit_summary)

# Create a bar plot
library(ggplot2)
ggplot(speed_limit_summary, aes(x = speed_limit, y = crash_count)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    theme_minimal() +
    labs(
        title = "Crash Frequency by Speed Limit",
        x = "Speed Limit",
        y = "Number of Crashes"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

head(crash_clean$Crash.Date.Time, 1)

nrow(crash_clean)

```

```{r}
library(sf)
library(tigris)
library(dplyr)
library(lubridate)
# Create dataframe with coordinates
crash_coords <- crash_clean %>% 
  select(Longitude, Latitude)

nrow(crash_coords)

crash_sf <- st_as_sf(crash_coords, coords = c("Longitude", "Latitude"), crs = 4326)

states <- states(cb = TRUE) %>%
  st_transform(crs = 4326)  # Transform states to WGS84

crash_with_states <- st_join(crash_sf, states["NAME"])

crash_states_df <- crash_with_states %>%
  mutate(row_num = row_number()) %>%
  st_drop_geometry() %>%
  select(NAME, row_num)

crash_clean_with_states <- crash_clean %>%
  mutate(row_num = row_number()) %>%
  left_join(crash_states_df, by = "row_num") %>%
  rename(state = NAME) %>%
  select(-row_num)

crash_maryland_injuries <- crash_clean_with_states %>% 
  mutate(date = mdy_hms(Crash.Date.Time)) %>%
  filter(year(date) == 2024,
#         state == "Maryland", 
         Injury.Severity != "NO APPARENT INJURY",
#         Injury.Severity != "POSSIBLE INJURY",
         Injury.Severity != "UNKNOWN",
         Vehicle.First.Impact.Location != "NON-COLLISION",
         Vehicle.First.Impact.Location != "VEHICLE NOT AT SCENE")

colnames(crash_maryland_injuries)

```

```{r}
colnames(crash_maryland_injuries)

unique(crash_maryland_injuries$Injury.Severity)

crash_maryland_injuries$Injury.Severity <- factor(crash_maryland_injuries$Injury.Severity, levels = c('SUSPECTED MINOR INJURY', 'SUSPECTED SERIOUS INJURY', 'FATAL INJURY'))

# First, create a contingency table
impact_injury_table <- table(crash_maryland_injuries$Vehicle.First.Impact.Location, 
                              crash_maryland_injuries$Injury.Severity)

# Convert to percentages if you want to see proportions
impact_injury_prop <- prop.table(impact_injury_table, margin = 1) * 100




# Create the plot with reordered bars
ggplot(data = as.data.frame.table(impact_injury_prop), 
       aes(x = reorder(Var1, -table(Var1)), y = Freq, fill = Var2)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_minimal() +
  labs(x = "Impact Location", 
       y = "Count",
       fill = "Injury Severity",
       title = "Relationship between Impact Location and Injury Severity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r}
# Chi-square test of independence
chisq_test <- chisq.test(table(crash_maryland_injuries$Vehicle.First.Impact.Location, 
                              crash_maryland_injuries$Injury.Severity))

chisq_test
# Effect size using Cramer's V
library(effectsize)
cramers_v <- cramers_v(crash_maryland_injuries$Vehicle.First.Impact.Location, 
                       crash_maryland_injuries$Injury.Severity)

# Standardized residuals to see which combinations occur more/less than expected
stdres <- chisq_test$stdres
stdres
```
```{r}
# Simple frequency table
impact_freq <- table(crash_maryland_injuries$Vehicle.First.Impact.Location)
# Sort in descending order
impact_freq_sorted <- sort(impact_freq, decreasing = TRUE)

# Chi-square goodness of fit test (tests against uniform distribution)
impact_chisq <- chisq.test(impact_freq)

# Visual representation
library(ggplot2)
impact_df <- data.frame(
  location = names(impact_freq),
  count = as.numeric(impact_freq)
)

ggplot(impact_df, aes(x = reorder(location, -count), y = count)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Impact Location",
       y = "Frequency",
       title = "Distribution of Vehicle Impact Locations")
```

```{r}
# Calculate weights (inverse of class frequencies)
weights <- 1 / table(crash_maryland_injuries$Vehicle.First.Impact.Location)
weights <- weights[crash_maryland_injuries$Vehicle.First.Impact.Location]

# Use weights in your analysis
weighted.chisq <- chisq.test(table(crash_maryland_injuries$Vehicle.First.Impact.Location, 
                                  crash_maryland_injuries$Injury.Severity),
                            weights)
weighted.chisq
```
```{r}



first_impact_severity <- crash_severity %>%
  group_by(Vehicle.First.Impact.Location) %>%
  summarise(
    avg_severity = mean(severity_score, na.rm = TRUE),
    crash_count = n()
  ) %>%  filter(Vehicle.First.Impact.Location != "UNKNOWN", Vehicle.First.Impact.Location != "VEHICLE NOT AT SCENE",crash_count > 10)

ggplot(first_impact_severity, aes(x = reorder(Vehicle.First.Impact.Location, -avg_severity), y = avg_severity)) +
  geom_bar(stat = "identity", fill = "black") +
  geom_text(aes(label = round(avg_severity, 2)), vjust = -0.5) +
  labs(
    title = "Average Crash Severity by Vehicle Impact Location",
    x = "Type of impact",
    y = "Average Severity Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



```{r}
library(mapview)


crash_maryland_sf <- st_as_sf(crash_maryland_injuries, 
                            coords = c("Longitude", "Latitude"), 
                            crs = 4326)

colnames(crash_maryland_sf)

severity_colors <- c("FATAL" = "red",
                    "NO APPARENT INJURY" = "green",
                    "POSSIBLE INJURY" = "blue",
                    "SUSPECTED MINOR INJURY" = "yellow",
                    "SUSPECTED SERIOUS INJURY" = "orange"
                    )




mapview(crash_maryland_sf,
        zcol = "Injury.Severity",
        col.regions = severity_colors,
        cex = 3,
        alpha = 0.8,
        legend = TRUE,
        popup = c("Injury.Severity"),
        layer.name = "Crash Severity")

mapview(crash_maryland_sf,
        zcol = "Vehicle.First.Impact.Location",
        col.regions = severity_colors,
        cex = 3,
        alpha = 0.8,
        legend = TRUE,
        popup = c("Injury.Severity"),
        layer.name = "Crash Severity")



```




## Analysis 2: Vehicle Types in Crashes

### Research Question
What car types are most commonly found in crashes, and is there a relationship with crash severity?

```{r vehicle-analysis}
# Analysis of crashes by vehicle type
vehicle_summary <- crash_clean %>%
  group_by(Vehicle.Body.Type) %>%
  summarise(
    crash_count = n(),
    severe_crashes = sum(Injury.Severity %in% c("FATAL INJURY", "SUSPECTED SERIOUS INJURY")),
    severity_rate = severe_crashes / crash_count
  )

# Visualization code will go here
```

## Analysis 3: Factors Affecting Crash Severity

### Research Question
What factors correlate most strongly with crash severity?

```{r time-analysis}
# Extract hour from Crash.Date.Time
crash_severity <- crash_severity %>%
  mutate(
    crash_hour = hour(as.POSIXct(Crash.Date.Time, format="%m/%d/%Y %I:%M:%S %p")),
    time_of_day = case_when(
      crash_hour >= 5 & crash_hour < 12 ~ "Morning",
      crash_hour >= 12 & crash_hour < 17 ~ "Afternoon",
      crash_hour >= 17 & crash_hour < 22 ~ "Evening",
      TRUE ~ "Night"
    )
  )

# Analyze severity by time of day
time_severity <- crash_severity %>%
  group_by(time_of_day) %>%
  summarise(
    avg_severity = mean(severity_score, na.rm = TRUE),
    crash_count = n()
  )

# Visualize time of day patterns
ggplot(time_severity, aes(x = time_of_day, y = avg_severity)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = round(avg_severity, 2)), vjust = -0.5) +
  labs(
    title = "Average Crash Severity by Time of Day",
    x = "Time of Day",
    y = "Average Severity Score"
  ) +
  theme_minimal()
```

```{r weather-analysis}
# Analyze severity by weather conditions
weather_severity <- crash_severity %>%
  group_by(Weather) %>%
  summarise(
    avg_severity = mean(severity_score, na.rm = TRUE),
    crash_count = n()
  ) %>%
  filter(!is.na(Weather), Weather != "", crash_count > 10)  # Remove sparse categories

# Visualize weather patterns
ggplot(weather_severity, aes(x = reorder(Weather, -avg_severity), y = avg_severity)) +
  geom_bar(stat = "identity", fill = "darkred") +
  geom_text(aes(label = round(avg_severity, 2)), vjust = -0.5) +
  labs(
    title = "Average Crash Severity by Weather Condition",
    x = "Weather Condition",
    y = "Average Severity Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r traffic-control-analysis}
# Analyze severity by traffic control presence
control_severity <- crash_severity %>%
  group_by(Traffic.Control) %>%
  summarise(
    avg_severity = mean(severity_score, na.rm = TRUE),
    crash_count = n()
  ) %>%
  filter(!is.na(Traffic.Control), Traffic.Control != "", crash_count > 10)

# Visualize traffic control patterns
ggplot(control_severity, aes(x = reorder(Traffic.Control, -avg_severity), y = avg_severity)) +
  geom_bar(stat = "identity", fill = "forestgreen") +
  geom_text(aes(label = round(avg_severity, 2)), vjust = -0.5) +
  labs(
    title = "Average Crash Severity by Traffic Control Type",
    x = "Traffic Control",
    y = "Average Severity Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r distraction-analysis}
# Analyze severity by driver distraction
distraction_severity <- crash_severity %>%
  group_by(Driver.Distracted.By) %>%
  summarise(
    avg_severity = mean(severity_score, na.rm = TRUE),
    crash_count = n()
  ) %>%
  filter(!is.na(Driver.Distracted.By), Driver.Distracted.By != "", crash_count > 10)

# Visualize distraction patterns
ggplot(distraction_severity, aes(x = reorder(Driver.Distracted.By, -avg_severity), y = avg_severity)) +
  geom_bar(stat = "identity", fill = "purple") +
  geom_text(aes(label = round(avg_severity, 2)), vjust = -0.5) +
  labs(
    title = "Average Crash Severity by Driver Distraction",
    x = "Distraction Type",
    y = "Average Severity Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r movement-analysis}
# Analyze severity by vehicle movement
movement_severity <- crash_severity %>%
  group_by(Vehicle.Movement) %>%
  summarise(
    avg_severity = mean(severity_score, na.rm = TRUE),
    crash_count = n()
  ) %>%
  filter(!is.na(Vehicle.Movement), Vehicle.Movement != "", crash_count > 10)

# Visualize movement patterns
ggplot(movement_severity, aes(x = reorder(Vehicle.Movement, -avg_severity), y = avg_severity)) +
  geom_bar(stat = "identity", fill = "orange") +
  geom_text(aes(label = round(avg_severity, 2)), vjust = -0.5) +
  labs(
    title = "Average Crash Severity by Vehicle Movement",
    x = "Vehicle Movement",
    y = "Average Severity Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r statistical-tests}
# Chi-square test for categorical variables
# Weather and Severity
weather_chi <- table(crash_severity$Weather, crash_severity$Injury.Severity)
weather_test <- chisq.test(weather_chi)

# Traffic Control and Severity
control_chi <- table(crash_severity$Traffic.Control, crash_severity$Injury.Severity)
control_test <- chisq.test(control_chi)

# Create a summary of statistical tests
test_results <- data.frame(
  Factor = c("Weather", "Traffic Control"),
  Chi_Square = c(weather_test$statistic, control_test$statistic),
  P_Value = c(weather_test$p.value, control_test$p.value)
)

# Print test results
print(test_results)

# Perform logistic regression for severe crashes
crash_severity$is_severe <- crash_severity$severity_score >= 3

# Create logistic regression model
severe_model <- glm(is_severe ~ 
                     Weather + 
                     time_of_day + 
                     Traffic.Control + 
                     Vehicle.Movement,
                   data = crash_severity,
                   family = "binomial")

# Print model summary
summary(severe_model)
```

```{r create-summary-table}
# Create a summary table of the most significant factors
factor_summary <- data.frame(
  Factor = c("Time of Day", "Weather", "Traffic Control", 
             "Driver Distraction", "Vehicle Movement"),
  Average_Severity = c(
    max(time_severity$avg_severity),
    max(weather_severity$avg_severity),
    max(control_severity$avg_severity),
    max(distraction_severity$avg_severity),
    max(movement_severity$avg_severity)
  ),
  Most_Severe_Category = c(
    time_severity$time_of_day[which.max(time_severity$avg_severity)],
    weather_severity$Weather[which.max(weather_severity$avg_severity)],
    control_severity$Traffic.Control[which.max(control_severity$avg_severity)],
    distraction_severity$Driver.Distracted.By[which.max(distraction_severity$avg_severity)],
    movement_severity$Vehicle.Movement[which.max(movement_severity$avg_severity)]
  )
)

# Print the summary table
print(factor_summary)
```

## Conclusions

[Your conclusions will go here after completing the analysis]

## Recommendations

[Your recommendations based on the findings will go here]

## References

[Add any references or data sources used]